Simple repression draft

### 2020-10-14 doc/resubmission/appendix_01_nonbursty.tex
more changes
fix appendix 1...again

\subsection{Thremodynamic models of gene regulation}
The first class of models we will explore are the so-called thermodynamic models
of gene regulation~\cite{Ackers1982}. The premise for these models is that we
imagine the steps leading to transcription --binding and unbinding of
transcription factors and of RNAP-- to occur in a timescale much faster than the
downstream processes --open complex formation and RNAP
escape~\cite{Browning2004}. What this allows us to do is to assume that the
steps leading to transcription can be modeled as being in quasi-equilibrium.
This is enormously advantageous since we can use the theoretical tools of 
equilibrium statisticla mechanics to model such process.

This means that we can enlist the possible microstates in which we can find the
promoter in order to calculate their probabilities based on quantities such as
molecular counts and interaction energies. We then assume that the gene
expression level is proportional to the probability of finding the promoter in
the transcriptionally active microstate, i.e.,
\begin{equation}
\text{average rate of transcription} = \sum_i r_i p_i,
\label{eq:rate_transcription_si}
\end{equation}
where $i$ labels the distinct states, $p_i$ is the probability of the ith state,
and $r_i$ is the rate of transcription of that state. As mentioned in the main
text, these models can only deal with the mean gene expression level. This is
because the probability space we consider is that of the state of the repressor,
rather than both the state of the repressor and the mRNA counts. Let's now work
through models 1 and 2 from Figure~\ref{fig1:means_cartoons}(B).

\subsubsection{The Two-State Equilibrium Model}
In this simplest model, depicted as (1) in Figure~\ref{fig1:means_cartoons}(B),
the promoter is idealized as existing in one of two states, either repressor
bound or repressor unbound. The rate of transcription
is assumed to be proportional to the fraction of time spent
in the repressor unbound state.
From the relative statistical weights listed in
Figure~\ref{fig1:means_cartoons}(B), the probability $p_U$ of being in the
unbound state is
\begin{equation}
p_U = \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1}.
\end{equation}
The mean rate of transcription is then given by $r p_U$, as assumed by
Eq.~\ref{eq:rate_transcription_si}. The mean number of mRNA is
set by the balance of average mRNA transcription and degradation rates,
so it follows that the mean mRNA level is given by
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
where $r$ is the transcription rate from the repressor unbound state, $\gamma$
is the mRNA degradation rate, $R$ is repressor copy number, $N_{NS}$ is the
number of nonspecific binding sites in the genome where repressors spend most of
their time when not bound to the operator (binding site), $\beta \equiv 1/k_BT$,
and $\Delta\varepsilon_R$ is the binding energy of a repressor to its operator
site. The derivation of this result can be found in~\cite{Phillips2015a}.

@@

In fact it was noted in~\cite{Chure2019} that this two-state model can be viewed
as the coarse-graining of any equilibrium promoter model in which no
transcriptionally active states have transcription factor bound, or put
differently, when there is no overlap between transcription factor bound states
and transcriptionally active states. We will see this explicitly in the 3-state
equilibrium model below, but perhaps surprising is that an analogous result
carries over even to the nonequilibrium models we consider later.

\subsubsection{The Three-State Equilibrium Model}
Compared to the previous model, here we fine-grain the repressor unbound state
into two separate states: empty, and RNAP bound as shown in (2) in
Figure~\ref{fig1:means_cartoons}(B). This picture was used in~\cite{Garcia2011a}
as we use it here, and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was
generalized to incorporate small-molecule inducers that bind the repressor. The
effect of this generalization is, roughly speaking, simply to rescale $R$ from
the total number of repressors to a smaller effective number of available
repressors which are unbound by inducers. We point out that the same
generalization can be incorporated quite easily into any of our models in
Figure~\ref{fig1:means_cartoons} by simply rescaling the repressor copy number
$R$ in the thermodynamic models, or equivalently $k_R^+$ in the kinetic models.

The mean mRNA copy number, as derived in Appendix~\ref{sec:non_bursty}
from a similar enumeration of states and weights as the previous model, is
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}
        },
\end{equation}
where the new variables are $\Delta\varepsilon_P$, the difference in RNAP
binding energy to its specific site (the promoter) relative to an average
nonspecific background site, and the RNAP copy number, $P$. The fold-change
again follows immediately as
\begin{align}
\text{fold-change}
&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}
        }
\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
        {\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
\\
&= \left(
1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}}
        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
\right)^{-1}
\\
&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
\end{align}
with $\Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
\frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$ as shown
in~\fig{fig1:means_cartoons}(B). Thus far, we see that the two thermodynamic
models, despite making different coarse-graining commitments, result in the same
functional form for the fold-change in mean gene expression.  We now explore how
kinetic models fare when faced with computing the same observable.

@@

% \subsubsection{Nonequilibrium Model Four - ``Active'' and ``Inactive'' States}
% Model 4 in Figure~\ref{fig1:means_cartoons}(C) is at the core of the theory
% in~\cite{Razo-Mejia2020}. At a glance the cartoon for this model may appear very
% similar to model 2, and mathematically it is, but the interpretation is rather
% different. In model 2, we interpreted the third state literally as an RNAP-bound
% promoter and modeled initiation of a transcript as triggering a promoter state
% change, making the assumption that an RNAP can only
% make one transcript at a time. In contrast, in the present model the promoter
% state does \textit{not} change when a transcript is initiated. So we no longer
% interpret these states as literally RNAP bound and unbound but instead as
% coarse-grained ``active'' and ``inactive'' states, the details of which we leave
% unspecified for now. We will comment more on this model below when we discuss
% Fano factors of models.

% Mathematically this model is very similar to models 1 and 2. Like model 1, the
% matrix $R$ describing transcription initiation is diagonal, namely
% \begin{equation}
% \mathbf{R} = \begin{pmatrix}
%                 0 & 0 & 0 \\ 
%                 0 & 0 & 0 \\ 
%                 0 & 0 & r
%         \end{pmatrix}.
% \end{equation}
% The master equation takes verbatim the same form as it did for model 1,
% Eq.~\ref{eq:2state_rep_cme}. Meanwhile the promoter transition
% matrix $\mathbf{K}$ is the same as Eq.~\ref{eq:3state_cme_matrices_pt1}
% from model 2, although we relabel the
% rate constants from $k_P^\pm$ to $k^\pm$ to reiterate that these are not simply
% RNAP binding and unbinding rates.

% Carrying out the algebra, the mean mRNA can be found to be
% \begin{equation}
% \langle m\rangle = \frac{r}{\gamma}
% \frac{k_R^- k^+}
% {k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
% \end{equation}
% and the fold-change readily follows,
% \begin{align}
% \text{fold-change}
% &=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
%         \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
% \\
% &=      \left(1 + \frac{k_R^+}{k_R^-}
%                 \left(1 + \frac{k^+}{k^-}\right)^{-1}
%         \right)^{-1},
% \end{align}
% from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}(C).

@@

Having derived an example chemical master equation we now focus on writing a
general matrix form for the kinetic models 1-4 shown in
Figure~\ref{fig1:means_cartoons}(C) in the main text. In each of these four
models, the natural mathematicization of their cartoons is as a chemical master
equation such as the coupled chemical master equations we derived for model 1,
which for completeness we reproduce here as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\label{eq:poisson_promoter_cme_appdx}
\end{split}
\end{align}
Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding the system with
$m$ mRNA molecules at time $t$ either in the repressor bound or unbound states,
respectively. $r$ is the probability per unit time that a transcript will be
initiated when the repressor is unbound, and $\gamma$ is the probability per
unit time for a given mRNA to be degraded. $k_R^-$ is the probability per unit
time that a bound repressor will unbind, while $k_R^+$ is the probability per
unit time that an unbound operator will become bound by a repressor. Assuming
mass action kinetics, $k_R^+$ is proportional to repressor copy number $R$.

Next consider the cartoon for nonequilibrium model 2 in
Figure~\ref{fig1:means_cartoons}(C). Now we must track probabilities $p_R$,
$p_P$, and $p_E$ for the repressor bound, empty, and polymerase bound states,
respectively. By analogy to Eq.~\ref{eq:poisson_promoter_cme_appdx}, the master
equation for model 2 can be written
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_E(m,t) =&\; 
  \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}.
\\
&
+ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U}
\\
\deriv{t}p_P(m,t) =&\; 
- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}.
\\
&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U}.
\label{eq:model2_cme_appdx}
\end{split}
\end{align}
$k_P^+$ and $k_P^-$ are defined in close analogy to $k_R^+$ and $k_R^-$, except
for RNAP binding and unbinding instead of repressor. Similarly $p_P(m,t)$ is
defined for the active (RNAP-bound) state exactly as are $p_R(m,t)$ and
$p_E(m,t)$ for the repressor bound and unbound states, respectively. The new
subtlety Eq.~\ref{eq:model2_cme_appdx} introduces compared to
Eq.~\ref{eq:poisson_promoter_cme_appdx} is that when mRNAs are produced, the
promoter state also changes. This is encoded by the terms involving $r$, the
last term in each of the equations for $p_E$ and $p_P$. The former accounts for
arrivals in the unbound state and the latter accounts for departures from the
RNAP-bound state.

@@

Although we derived Eq.~\ref{eq:generic_cme_appdx} for the particular case of
kinetic model 2 in Figure~\ref{fig1:means_cartoons}, in fact the chemical master
equations for all of the kinetic models in Figure~\ref{fig1:means_cartoons}
except for model 5 can be cast in this form. (We treat model 5 separately in
Appendix~\ref{sec:gen_fcn_appdx}.) Model 3 introduces no new subtleties beyond
model 2 and Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply
with different matrices of dimension four instead of three. Models 1 and 4 are
both handled by a special case of Eq.~\ref{eq:generic_cme_appdx} where
$\mathbf{R_D} = \mathbf{R_A} \equiv \mathbf{R}$, since in these two models
transcription initiation events do not change promoter state.

Recalling that our goal in this section is to derive expressions for mean mRNA
and Fano factor for nonequilibrium models 1 through four in
Figure~\ref{fig1:means_cartoons}, and since all four of these models are
described by Eq.~\ref{eq:generic_cme_appdx}, we can save substantial effort by
deriving general formulas for mean mRNA and Fano factor from
Eq.~\ref{eq:generic_cme_appdx} once and for all. Then for each model we can
simply plug in the appropriate matrices for $\mathbf{K}$, $\mathbf{R_D}$, and
$\mathbf{R_A}$ and carry out the remaining algebra.

@@

\subsection{Kinetic Model One - Poisson Promoter}
\subsubsection{Mean mRNA}
For kinetic model 1 in Figure~\ref{fig1:means_cartoons}, we have already shown
the full master equation in Eq.~\ref{eq:poisson_promoter_cme_appdx}, but for
completeness we reprint it again as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\end{split}
\label{eq:2state_rep_cme}
\end{align}
This is a direct transcription of the states and rates in
Figure~\ref{fig1:means_cartoons}(C). This may be converted to the matrix form of
the master equation shown in Eq.~\ref{eq:generic_cme_appdx} with matrices
\begin{equation}
\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_U(m) \end{pmatrix},\
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\mathbf{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\end{equation}
where $\mathbf{R_A}$ and $\mathbf{R_D}$ are equal, so we drop the subscript and
denote both simply by $\mathbf{R}$. Since our interest is only in steady-state
we dropped the time dependence as well.

First we need $\langle\vec{m}^0\rangle$. Label its components as $p_R$ and
$p_U$, the probabilities of finding the system in either promoter state, and
note that only $p_U$ survives multiplication by $\mathbf{R}$, since
\begin{equation}
\mathbf{R} \langle\vec{m}^0\rangle
= \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix}
    \begin{pmatrix} p_R \\ p_U \end{pmatrix}
= \begin{pmatrix} 0 \\ r p_U \end{pmatrix},
\label{eq:Rm0_model1_appdx}
\end{equation}
so we need not bother finding $p_R$. Then we have
\begin{equation}
(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle
= \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix}
    \begin{pmatrix} p_R \\ p_U \end{pmatrix} = 0.
\label{eq:model1_m0_giver_appdx}
\end{equation}
As mentioned earlier in Section~\ref{sec:m0_appdx}, the two rows are linearly
dependent, so taking only the first row and using normalization to set $p_R =
1-p_U$ gives
\begin{equation}
-k_R^- (1-p_U) + k_R^+ p_U = 0,
\end{equation}
which is easily solved to find
\begin{equation}
p_U = \frac{k_R^-}{k_R^- + k_R^+}.
\end{equation}
Substituting this into Eq.~\ref{eq:Rm0_model1_appdx}, and the result of that
into Eq.~\ref{eq:mean_m_appdx_ref}, we have
\begin{equation}
\langle{m}\rangle = \frac{r}{\gamma} \frac{k_R^-}{k_R^- + k_R^+}.
\end{equation}

\subsubsection{Fold-Change}
$k_R^+$ is proportional to the repressor copy number, so in computing
fold-change, absence of repressor corresponds to $k_R^+\rightarrow0$. Therefore
fold-change in this model is simply
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
matching the master curve of~\fig{fig1:means_cartoons}(D) with $\rho=1$.

@@

\subsection{Kinetic Model Two - RNAP Bound and Unbound States}

>>>>

\subsection{Thremodynamic models of gene regulation}
The first class of models we will explore are the so-called thermodynamic models
of gene regulation~\cite{Ackers1982}. The premise for these models is that we
imagine the steps leading to transcription --binding and unbinding of
transcription factors and of RNAP-- to occur in a timescale much faster than the
downstream processes --open complex formation and RNAP
escape~\cite{Browning2004}. What this allows us to do is to assume that the
steps leading to transcription can be modeled as being in quasi-equilibrium.
This is enormously advantageous since we can use the theoretical tools of 
equilibrium statistical mechanics to model such process.

This means that we can enlist the possible microstates in which we can find the
promoter in order to calculate their probabilities based on quantities such as
molecular counts and interaction energies. We then assume that the gene
expression level is proportional to the probability of finding the promoter in
the transcriptionally active microstate. As mentioned in the main text, these
models can only deal with the mean gene expression level. This is because the
probability space we consider is that of the state of the repressor, rather than
both the state of the repressor and the mRNA counts. Let's now work through
models 1 and 2 from Figure~\ref{fig1:means_cartoons}(B).

\subsubsection{The Two-State Equilibrium Model}
In this simplest model, depicted as (1) in Figure~\ref{fig1:means_cartoons}(B),
the promoter is idealized as existing in one of two states, either repressor
bound or repressor unbound. The rate of transcription
is assumed to be proportional to the fraction of time spent
in the repressor unbound state.
From the relative statistical weights listed in
Figure~\ref{fig1:means_cartoons}, the probability $p_U$ of being in the
unbound state is
\begin{equation}
p_U = \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1}.
\end{equation}
The mean rate of transcription is then given by $r p_U$, as assumed by
Eq.~\ref{eq:transcrip_prop_pbound}. The mean number of mRNA is
set by the balance of average mRNA transcription and degradation rates,
so it follows that the mean mRNA level is given by
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
where $r$ is the transcription rate from the repressor unbound state, $\gamma$
is the mRNA degradation rate, $R$ is repressor copy number, $N_{NS}$ is the
number of nonspecific binding sites in the genome where repressors spend most
of their time when not bound to the operator, $\beta \equiv 1/k_BT$, and
$\Delta\varepsilon_R$ is the binding energy of a repressor to its operator site.

\paragraph{Fold-change}

@@

In fact it was noted in~\cite{Chure2019} that this two-state model can be viewed
as the coarse-graining of any equilibrium promoter model in which no
transcriptionally active states have transcription factor bound, or put
differently, when there is no overlap between
transcription factor bound states and transcriptionally
active states. We will see this explicitly in the 3-state equilibrium model
below, but perhaps surprising is that an analogous result carries over even to
the kinetic models we consider later.

\subsubsection{The Three-State Equilibrium Model}
Compared to the previous model, here we fine-grain the repressor unbound state
into two separate states: empty, and RNAP bound as shown in (2) in
Figure~\ref{fig1:means_cartoons}(B). This picture was used in~\cite{Garcia2011a}
as we use it here, and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was
generalized to incorporate small-molecule inducers that bind the repressor. 
The effect of this generalization is, roughly speaking,
simply to rescale $R$ from the total number of repressors to a smaller
effective number of available repressors which are unbound by inducers.
We point out that the same generalization can be incorporated quite easily
into any of our models in Figure~\ref{fig1:means_cartoons} by simply
rescaling the repressor copy number $R$ in the equilibrium models, or
equivalently $k_R^+$ in the kinetic models.

The mean mRNA copy number, as derived in Appendix~\ref{sec:non_bursty}
from a similar enumeration of states and weights as the previous model, is
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}
        },
\end{equation}
where the new variables are $\Delta\varepsilon_P$, the difference in RNAP
binding energy to its specific site (the promoter) relative to an average
nonspecific background site, and the RNAP copy number, $P$.

\paragraph{Fold-change}
The fold-change again follows immediately as
\begin{align}
\text{fold-change}
&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}
        }
\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
        {\frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
\\
&= \left(
1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}}
        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
\right)^{-1}
\\
&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
\end{align}
with $\Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
\frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$ as shown
in~\fig{fig1:means_cartoons}(B). Thus far, we see that the two thermodynamic
models, despite making different coarse-graining commitments, result in the same
functional form for the fold-change in mean gene expression.  We now explore how
kinetic models fare when faced with computing the same observable.

@@

@@

Having derived an example chemical master equation we now focus on writing a
general matrix form for the kinetic models 1-4 shown in
Figure~\ref{fig1:means_cartoons}(C) in the main text. In each of these four
models, the natural mathematicization of their cartoons is as a chemical master
equation such as Eq.~\ref{eq:2state_rep_cme} for model 1, which for
completeness we reproduce here as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\label{eq:poisson_promoter_cme_appdx}
\end{split}
\end{align}
Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding the system with
$m$ mRNA molecules at time $t$ either in the repressor bound or unbound states,
respectively. $r$ is the probability per unit time that a transcript will be
initiated when the repressor is unbound, and $\gamma$ is the probability per
unit time for a given mRNA to be degraded. $k_R^-$ is the probability per unit
time that a bound repressor will unbind, while $k_R^+$ is the probability per
unit time that an unbound operator will become bound by a repressor. Assuming
mass action kinetics, $k_R^+$ is proportional to repressor copy number $R$.

Next consider the cartoon for kinetic model 2 in
Figure~\ref{fig1:means_cartoons}(C). Now we must track probabilities $p_R$,
$p_P$, and $p_E$ for the repressor bound, empty, and polymerase bound states,
respectively. By analogy to Eq.~\ref{eq:poisson_promoter_cme_appdx}, the master
equation for model 2 can be written
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_E(m,t) =&\; 
  \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_E(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_E(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_E(m,t)}^{m \rightarrow m - 1}.
\\
&
+ \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
- \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{rp_P(m-1,t)}^{m-1 \rightarrow m,\;A \rightarrow U}
\\
\deriv{t}p_P(m,t) =&\; 
- \overbrace{k_P^- p_P(m,t)}^{A \rightarrow U}
+ \overbrace{k_P^+ p_E(m,t)}^{U \rightarrow A}
+ \overbrace{(m+1)\gamma p_P(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_P(m,t)}^{m \rightarrow m - 1}.
\\
&- \overbrace{rp_P(m,t)}^{m \rightarrow m + 1,\;A \rightarrow U}.
\label{eq:model2_cme_appdx}
\end{split}
\end{align}
$k_P^+$ and $k_P^-$ are defined in close analogy to $k_R^+$ and $k_R^-$, except
for RNAP binding and unbinding instead of repressor. Similarly $p_P(m,t)$ is
defined for the active (RNAP-bound) state exactly as are $p_R(m,t)$ and
$p_E(m,t)$ for the repressor bound and unbound states, respectively. The new
subtlety Eq.~\ref{eq:model2_cme_appdx} introduces compared to
Eq.~\ref{eq:poisson_promoter_cme_appdx} is that when mRNAs are produced, the
promoter state also changes. This is encoded by the terms involving $r$, the
last term in each of the equations for $p_E$ and $p_P$. The former accounts for
arrivals in the unbound state and the latter accounts for departures from the
RNAP-bound state.

@@

Although we derived Eq.~\ref{eq:generic_cme_appdx} for the particular case of
kinetic model 2 in Figure~\ref{fig1:means_cartoons}, in fact the
chemical master equations for all of the kinetic models in
Figure~\ref{fig1:means_cartoons} except for model 5 can be cast in this form.
(We treat model 5 separately in Appendix~\ref{sec:gen_fcn_appdx}.) Model
3 introduces no new subtleties beyond model 2 and
Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply with different
matrices of dimension four instead of three. Models 1 and 4 are both
handled by Eq.~\ref{eq:2state_rep_cme} in the main text, which is just
Eq.~\ref{eq:generic_cme_appdx} except in the special case of $\mathbf{R_D} =
\mathbf{R_A} \equiv \mathbf{R}$, since in these two models transcription
initiation events do not change promoter state.

Recalling that our goal in this section is to derive expressions for mean mRNA
and Fano factor for kinetic models 1 through four in
Figure~\ref{fig1:means_cartoons}, and since all four of these models are
described by Eq.~\ref{eq:generic_cme_appdx}, we can save substantial effort by
deriving general formulas for mean mRNA and Fano factor from
Eq.~\ref{eq:generic_cme_appdx} once and for all. Then for each model we can
simply plug in the appropriate matrices for $\mathbf{K}$, $\mathbf{R_D}$, and
$\mathbf{R_A}$ and carry out the remaining algebra.

@@

\subsection{Kinetic Model One - Poisson Promoter}

This first model (model 1 in Figure~\ref{fig1:means_cartoons}) is the one that
we used in the previous section to motivate the derivation of the chemical
master equation. Eq.~\ref{eq:poisson_promoter_cme_appdx} shows the two coupled
master equations, but for completeness we reprint it again as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\end{split}
\label{eq:model1_kinetic_cme}
\end{align}
This is a direct transcription of the states and rates in
Figure~\ref{fig1:means_cartoons}. This may be converted to the matrix form of
the master equation shown in Eq.~\ref{eq:generic_cme_appdx} with matrices
\begin{equation}
\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_U(m) \end{pmatrix},\
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\mathbf{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\end{equation}
where $\mathbf{R_A}$ and $\mathbf{R_D}$ are equal, so we drop the subscript and
denote both simply by $\mathbf{R}$. Since our interest is only in steady-state
we dropped the time dependence as well.

\subsubsection{Mean mRNA}
To compute the mean mRNA we need $\langle\vec{m}^0\rangle$. Label its components
as $p_R$ and $p_U$, the probabilities of finding the system in either promoter
state, and note that only $p_U$ survives multiplication by $\mathbf{R}$, since
\begin{equation}
\mathbf{R} \langle\vec{m}^0\rangle
= \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix}
    \begin{pmatrix} p_R \\ p_U \end{pmatrix}
= \begin{pmatrix} 0 \\ r p_U \end{pmatrix},
\label{eq:Rm0_model1_appdx}
\end{equation}
so we need not bother finding $p_R$. Then we have
\begin{equation}
(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A}) \langle\vec{m}^0\rangle
= \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix}
    \begin{pmatrix} p_R \\ p_U \end{pmatrix} = 0.
\label{eq:model1_m0_giver_appdx}
\end{equation}
As mentioned earlier in Section~\ref{sec:m0_appdx}, the two rows are linearly
dependent, so taking only the first row and using normalization to set $p_R =
1-p_U$ gives
\begin{equation}
-k_R^- (1-p_U) + k_R^+ p_U = 0,
\end{equation}
which is easily solved to find
\begin{equation}
p_U = \frac{k_R^-}{k_R^- + k_R^+}.
\end{equation}
Substituting this into Eq.~\ref{eq:Rm0_model1_appdx}, and the result of that
into Eq.~\ref{eq:mean_m_appdx_ref}, we have
\begin{equation}
\langle{m}\rangle = \frac{r}{\gamma} \frac{k_R^-}{k_R^- + k_R^+}.
\end{equation}

\subsubsection{Fold-change}
The fold-change for this model is easily computed since the unregulated promoter
has a mean mRNA $\langle m \rangle = r / \gamma$, resulting in a fold-change of
the form
\begin{equation}
\text{fold-change} = \frac{k_R^-}{k_R^- + k_R^+}.
\end{equation}
This can be rewritten as a Fermi function of the form
\begin{equation}
\text{fold-change} = \frac{1}{1 + \exp[\log(k_R^+ / k_R^-)]},
\end{equation}
giving $\rho = 1$ as indicated in Figure~\ref{fig1:means_cartoons}.

@@

\subsection{Kinetic Model Two - RNAP Bound and Unbound States with RNAP escape}
